Radio Crickets: Chirping Jets from Black Hole Binaries Entering their Gravitational Wave Inspiral

### 2015-07-21 Paper/paper.tex
Add acknowledgements

\documentclass[a4paper,fleqn,usenatbib]{mnras}

% MNRAS is set in Times font. If you don't have this installed (most LaTeX
% installations will be fine) or prefer the old Computer Modern fonts, comment
% out the following line
%% \usepackage{newtxtext,newtxmath}

% Depending on your LaTeX fonts installation, you might get better results with one of these:
%% \usepackage{mathptmx}
%% \usepackage{txfonts}

% Use vector fonts, so it zooms properly in on-screen viewing software
% Don't change these lines unless you know what you are doing
\usepackage[T1]{fontenc}
\usepackage{ae,aecompl}

%%%%% AUTHORS - PLACE YOUR OWN PACKAGES HERE %%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%% AUTHORS - PLACE YOUR OWN COMMANDS HERE %%%%%

% Please keep new commands to a minimum, and use \newcommand not \def to avoid
% overwriting existing commands. Example:
%\newcommand{\pcm}{\,cm$^{-2}$}	% per cm-squared

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%% TITLE PAGE %%%%%%%%%%%%%%%%%%%

@@

\author[Kulkarni and Loeb]{
Girish Kulkarni,$^{1}$\thanks{E-mail: kulkarni@ast.cam.ac.uk}
and Abraham Loeb$^{2}$\thanks{E-mail: aloeb@cfa.harvard.edu}
\\
$^{1}$Institute of Astronomy and Kavli
  Institute of Cosmology, University of Cambridge, Madingley Road,
  Cambridge CB3 0HA, UK\\ 
$^{2}$Institute for Theory \& Computation,
  Harvard University, 60 Garden Street, Cambridge, MA 02138, USA\\ 
}

% These dates will be filled out by the publisher
\date{Accepted ---. Received ---; in original form ---}

% Enter the current year, for the copyright statements etc.
\pubyear{2015}

% Don't change these lines
\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

% Abstract of the paper
\begin{abstract}
  We study a novel electromagnetic signature of supermassive black
  hole binaries whose inspiral stars being dominated by gravitational
  wave (GW) emission. Combined GW and electromagnetic observations of
  compact binary mergers would allow to use them as standard sirens
  for cosmology and a testing ground of general relativity.  If one of
  the binary members produces a radio jet, the helical chirping
  morphology of the jet can be used to infer the binary parameters.
  For binaries with mass $10^7$--$10^{10}$ M$_\odot$ at redshifts
  $z<0.5$, the jet precesses along a biconical surface due to the
  binary's orbital motion.  When the binary enters the GW phase of its
  evolution, the opening angle widens, the jet exhibits milliarcsecond
  scale wiggles and the conical surface of jet precession is twisted
  due to the superluminal motion.  Monitoring these features in
  current and archival data will place a lower limit on sources that
  could be detected by eLISA and Pulsar Timing Arrays.  In the future,
  microarcsecond interferometry with the Square Kilometer Array will
  increase the potential usefulness of this technique.
\end{abstract}

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
keyword1 -- keyword2 -- keyword3
\end{keywords}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%% BODY OF PAPER %%%%%%%%%%%%%%%%%%

@@

% Any more references to add here?
General relativisitic (GR) numerical simulations can predict the
binary's evolution in its final stage up to coalescence
\citep{2005PhRvL..95l1101P, 2006PhRvL..96k1102B, 2006PhRvL..96k1101C}.
This has stimulated interest in predicting related electromagnetic
signals of black hole coalescence.  Effects considered in the
literature include afterglows due to infall of gas on the remnant
black hole \citep{2005ApJ...622L..93M}, electromagnetic variability in
the circumbinary disk due to shocks induced by the sudden mass loss of
the binary due to GW emission at coalescence
\citep{2007APS..APR.S1010B}, flares from shocked remnants of accretion
disks around recoiled remnant black hole \citep{2007PhRvL..99d1103L,
  2008ApJ...682..758S, 2009CQGra..26i4032H}, quasi-periodic
variability in gas accretion in the early stages of black hole merger
\citep{2008ApJ...672...83M, 2009CQGra..26i4032H, 2012MNRAS.427.2680K},
evolution in the profiles of broad emission lines shortly before
merger \citep{2013MNRAS.432.1468M}, characteristic evolution in
thermal emission due to circumbinary accretion
\citep{2015MNRAS.446L..36F}, brightening of accretion disk due to
viscous dissipation of gravitational waves
\citep{2008PhRvL.101d1101K}, and prompt tidal disruption of stars by
the recoiled remnant \citep{2011MNRAS.412...75S}.

% Replace `LISA' with `eLISA` or `NGO'?  Mention pulsar timing arrays?
In addition to testing GR in the strong field limit, observations of
electromagnetic counterparts will help localize sources of GWs for
future detectors such as the Evolved Laser Interferometer Space
Antenna (eLISA)\footnote{\url{https://www.elisascience.org}}, which would be
capable of observing the peak GW luminosity for BH coalescence of
$\sim 10^{57}$ erg$/$s up to large cosmological distances
\citep{2003CQGra..20S..65H, 2013CQGra..30x4009S}.  Simultaneous
detection of GW and electromagnetic signals from coalescing binaries
will yield rich scientific pay-offs \citep{2003CQGra..20S..65H,
  2005ApJ...629...15H}.  For example, by a redshift measurement of
eLISA sources could constraint cosmological parameters.  Measurement
of the black hole masses and spins from the GW signal will constrain
accretion physics during coalescence.  But even before eLISA is
launched, a search of electromagnetic signatures may point to the
expected abundance of GW sources.

@@

We study a BH binary formed following a gas-rich merger of two
galaxies as a result of the processes described above.  We consider
black holes with masses $M_1$ and $M_2$ in a circular orbit of radius
$a$ around each other.  Let $M=M_1+M_2$.  For simplicity, we consider
an equal-mass binary ($M_1=M_2$) on a circular orbit. We will consider
unequal black holes masses and elliptical orbits below.  The orbital
speed of each black hole is given by \citep{2010PhRvD..81d7503L}
\begin{equation}
  v = \frac{1}{2}\left(\frac{GM}{a}\right)^{1/2}\!\!\!\! = 5.8\times 10^3\, \mathrm{km~s}^{-1}M_8^{1/2}a_{16}^{-1/2},
  \label{eqn:vorb}
\end{equation}
where $a_{16}\equiv (a/10^{16}$ cm$)$ and $M_8\equiv (M/10^8$
M$_\odot)$ are the binary separation and mass,
respectively.\footnote{Note that $10^{16} \mathrm{cm}\approx 3.4\times
  10^{-3} \mathrm{pc}$.  The Schwarzchild radius of a $10^8$~M$_\odot$
  black hole is $9.6\times 10^{-6}$~pc, or $\sim 3\times 10^{13}$~cm.}
The orbital period is
\begin{equation}
  P = 2\pi\left(\frac{GM}{a^3}\right)^{-1/2}\!\!\!\! = 1.72\, \mathrm{yr}\, a_{16}^{3/2} M_8^{-1/2}.
\end{equation}
% Mention typical size of accretion disks and typical ISCO radius
% above?

@@

\begin{figure*}
\begin{center}
  \begin{tabular}{cc}
    \includegraphics*[scale=0.45]{{jet_i40_beta0.90_mdot1.00_full}.pdf} &
    \includegraphics*[scale=0.45]{jet_check_zoom1.pdf}
  \end{tabular}
\end{center}
\caption{The left panel shows forward and backward jets for an
  equal-mass binary with total mass $M=10^{10}$~M$_\odot$.  We have
  assumed $i=40^\circ$, $\beta=0.9$, and $d = 100$~Mpc.  The jets are
  twisted because of the apparant superluminal motion.  Also evident
  is the opening of the jet and the radio jet chirp.  Right panel
  shows normalised brightness of the inner 30~mas region of the
  forward jet.}
\label{fig:jet_mdot1.0}
\end{figure*}

@@

Finally, we assume that the emission from each jet particle is
optically thin at the frequency of observation with a spectrum in the
rest frame of the particle given by $P(\nu)\propto \nu^{-\alpha}$
\citep{1982ApJ...262..478G}.  The observed flux density $S(\nu)$ in erg
cm$^{-2}$ Hz$^{-1}$ is given by
\begin{equation}
  S(\nu)=S_r(\nu)D^{3+\alpha},
\end{equation}
where $D$ is the Doppler shift parameter
$[\gamma(1-\beta\cos\phi)]^{-1}$ and $S_r$ is the rest frame flux
density.  Further to consider the effects of jet evolution we assume
that $S_r$ decreases with time according to a simple power law
\begin{equation}
  S_r\propto t^{-\delta}.
\end{equation}
We assume $\delta=1$ and $\alpha=1$, hereafter. 
% Need to justify parameter values.
% Comment on relativistic corrections

@@

As the binary evolves, its increasing orbital period and speed, and
the apparant superluminal motion effect, together result in a jet
morphology with an increasing opening angle and increasingly rapid
periodic variation together with a twisted shape.
Figure~\ref{fig:jet_mdot1.0} shows these three features for an
equal-mass binary with total mass $M=10^{10}$~M$_\odot$.  For
concreteness, we have assumed $i=40^\circ$ and $\beta=0.9$.  The
angular diameter distance to the binary is assumed to be $d =
100$~Mpc, which corresponds to redshift $z\sim 0.3$.  The forward jet
streched on the sky while the backward jet is compressed due to the
superluminal motion.  The right panel of Figure~\ref{fig:jet_mdot1.0}
zooms into the inner region of the forward jet and shows that the
binary creates wiggles in the jet morphology of the order of a
milli-arcsecond.  The duration of the binary evolution captured by
Figure~\ref{fig:jet_mdot1.0} is 100 yr.

% Add discussion about the chirp of the signal

@@

% Give redshift numbers
These milli-arcsecond features can be resolved using long baseline
radio interferometry \citep{2001ChJAA...1..236Q}.  At higher
redshifts, jets are fainter and the wiggles become more difficult to
resolve.  However, it might be possible to resolve binaries with very
unequal mass ratios, eccentric orbits, or suitable inclinations.  The
future Square Kilometer Array (SKA) should make microarcsecond
resolution possible, which could increase the detection rate of these
objects \citep{2014arXiv1412.5971P}.

The jet morphology acts as an electromagnetic counterpart for chirping
gravitational wave sources.  BH binaries are expected to be the
brightest GW sources in the universe.  Current ground based GW
detectors such as LIGO are sensitive to low-mass BH binaries at high
redshift, which would be difficult to detect with this method.
However, future detectors such as eLISA are sensitvie to GW emission
from $10^7$--$10^9$ M$_\odot$ BHs \citep{2013CQGra..30x4009S}.  Pulsar
Timing Arrays are sensitive to even higher binary masses.  Monitoring
of jet morphologies with features outlined in this paper will put a
lower bound on the expected abundance of bright GW sources for this
experiments. Moreover, combined GW and electromagnetic observations of
compact binary mergers should enable detailed studies of astrophysical
processes in the strong-field gravity regime
\citep{2005ApJ...629...15H}.

% Mention old observations which show similar features
% There is a todo in a PS email from Avi

@@

\bibliographystyle{mnras}
\bibliography{refs} % if your bibtex file is called example.bib

% Don't change these lines
\bsp	% typesetting comment
\label{lastpage}
\end{document}

>>>>

\documentclass[a4paper,fleqn,usenatbib]{mnras}
\usepackage[T1]{fontenc}
\usepackage{ae,aecompl}

@@

@@

\author[Kulkarni and Loeb]{
Girish Kulkarni$^{1}$\thanks{E-mail: kulkarni@ast.cam.ac.uk}
and Abraham Loeb$^{2}$\thanks{E-mail: aloeb@cfa.harvard.edu}
\\
$^{1}$Institute of Astronomy and Kavli
  Institute of Cosmology, University of Cambridge, Madingley Road,
  Cambridge CB3 0HA, UK\\ 
$^{2}$Institute for Theory \& Computation,
  Harvard University, 60 Garden Street, Cambridge, MA 02138, USA\\ 
}

\date{Accepted ---. Received ---; in original form ---}

\pubyear{2015}

\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

\begin{abstract}
  We study a novel electromagnetic signature of supermassive black
  hole binaries whose inspiral stars being dominated by gravitational
  wave (GW) emission. Combined GW and electromagnetic observations of
  compact binary mergers would allow to use them as standard sirens
  for cosmology and a testing ground of general relativity.  If one of
  the binary members produces a radio jet, the helical chirping
  morphology of the jet can be used to infer the binary parameters.
  For binaries with mass $10^7$--$10^{10}$ M$_\odot$ at redshifts
  $z<0.5$, the jet precesses along a biconical surface due to the
  binary's orbital motion.  When the binary enters the GW phase of its
  evolution, the opening angle widens, the jet exhibits milliarcsecond
  scale wiggles and the conical surface of jet precession is twisted
  due to the superluminal motion.  Monitoring these features in
  current and archival data will place a lower limit on sources that
  could be detected by eLISA and Pulsar Timing Arrays.  In the future,
  microarcsecond interferometry with the Square Kilometer Array will
  increase the potential usefulness of this technique.
\end{abstract}

\begin{keywords}
keyword1 -- keyword2 -- keyword3
\end{keywords}

@@

General relativisitic (GR) numerical simulations can predict the
binary's evolution in its final stage up to coalescence
\citep{2005PhRvL..95l1101P, 2006PhRvL..96k1102B, 2006PhRvL..96k1101C}.
This has stimulated interest in predicting related electromagnetic
signals of black hole coalescence.  Effects considered in the
literature include afterglows due to infall of gas on the remnant
black hole \citep{2005ApJ...622L..93M}, electromagnetic variability in
the circumbinary disk due to shocks induced by the sudden mass loss of
the binary due to GW emission at coalescence
\citep{2007APS..APR.S1010B}, flares from shocked remnants of accretion
disks around recoiled remnant black hole \citep{2007PhRvL..99d1103L,
  2008ApJ...682..758S, 2009CQGra..26i4032H}, quasi-periodic
variability in gas accretion in the early stages of black hole merger
\citep{2008ApJ...672...83M, 2009CQGra..26i4032H, 2012MNRAS.427.2680K},
evolution in the profiles of broad emission lines shortly before
merger \citep{2013MNRAS.432.1468M}, characteristic evolution in
thermal emission due to circumbinary accretion
\citep{2015MNRAS.446L..36F}, brightening of accretion disk due to
viscous dissipation of gravitational waves
\citep{2008PhRvL.101d1101K}, and prompt tidal disruption of stars by
the recoiled remnant \citep{2011MNRAS.412...75S}.

In addition to testing GR in the strong field limit, observations of
electromagnetic counterparts will help localize sources of GWs for
future detectors such as the Evolved Laser Interferometer Space
Antenna (eLISA)\footnote{\url{https://www.elisascience.org}}, which would be
capable of observing the peak GW luminosity for BH coalescence of
$\sim 10^{57}$ erg$/$s up to large cosmological distances
\citep{2003CQGra..20S..65H, 2013CQGra..30x4009S}.  Simultaneous
detection of GW and electromagnetic signals from coalescing binaries
will yield rich scientific pay-offs \citep{2003CQGra..20S..65H,
  2005ApJ...629...15H}.  For example, by a redshift measurement of
eLISA sources could constraint cosmological parameters.  Measurement
of the black hole masses and spins from the GW signal will constrain
accretion physics during coalescence.  But even before eLISA is
launched, a search of electromagnetic signatures may point to the
expected abundance of GW sources.

@@

We study a BH binary formed following a gas-rich merger of two
galaxies as a result of the processes described above.  We consider
black holes with masses $M_1$ and $M_2$ in a circular orbit of radius
$a$ around each other.  Let $M=M_1+M_2$.  For simplicity, we consider
an equal-mass binary ($M_1=M_2$) on a circular orbit. We will consider
unequal black holes masses and elliptical orbits below.  The orbital
speed of each black hole is given by \citep{2010PhRvD..81d7503L}
\begin{equation}
  v = \frac{1}{2}\left(\frac{GM}{a}\right)^{1/2}\!\!\!\! = 5.8\times 10^3\, \mathrm{km~s}^{-1}M_8^{1/2}a_{16}^{-1/2},
  \label{eqn:vorb}
\end{equation}
where $a_{16}\equiv (a/10^{16}$ cm$)$ and $M_8\equiv (M/10^8$
M$_\odot)$ are the binary separation and mass,
respectively.\footnote{Note that $10^{16} \mathrm{cm}\approx 3.4\times
  10^{-3} \mathrm{pc}$.  The Schwarzchild radius of a $10^8$~M$_\odot$
  black hole is $9.6\times 10^{-6}$~pc, or $\sim 3\times 10^{13}$~cm.}
The orbital period is
\begin{equation}
  P = 2\pi\left(\frac{GM}{a^3}\right)^{-1/2}\!\!\!\! = 1.72\, \mathrm{yr}\, a_{16}^{3/2} M_8^{-1/2}.
\end{equation}

@@

\begin{figure*}
\begin{center}
  \begin{tabular}{cc}
    % jet3d_check2.py (case 0) 
    \includegraphics*[scale=0.45]{{jet_i40_beta0.90_mdot1.00_full}.pdf} &
    % jet3d_image_fulljet.py (case 1) 
    \includegraphics*[scale=0.45]{jet_check_zoom1.pdf}
  \end{tabular}
\end{center}
\caption{The left panel shows forward and backward jets for an
  equal-mass binary with total mass $M=10^{10}$~M$_\odot$.  We have
  assumed $i=40^\circ$, $\beta=0.9$, and $d = 100$~Mpc.  The jets are
  twisted because of the apparant superluminal motion.  Also evident
  is the opening of the jet and the radio jet chirp.  Right panel
  shows normalised brightness of the inner 30~mas region of the
  forward jet.}
\label{fig:jet_mdot1.0}
\end{figure*}

@@

Finally, we assume that the emission from each jet particle is
optically thin at the frequency of observation with a spectrum in the
rest frame of the particle given by $P(\nu)\propto \nu^{-\alpha}$
\citep{1982ApJ...262..478G}.  The observed flux density $S(\nu)$ in erg
cm$^{-2}$ Hz$^{-1}$ is given by
\begin{equation}
  S(\nu)=S_r(\nu)D^{3+\alpha},
\end{equation}
where $D$ is the Doppler shift parameter
$[\gamma(1-\beta\cos\phi)]^{-1}$ and $S_r$ is the rest frame flux
density.  Further to consider the effects of jet evolution we assume
that $S_r$ decreases with time according to a simple power law
\begin{equation}
  S_r\propto t^{-\delta}.
\end{equation}
We assume $\delta=1$ and $\alpha=1$, hereafter
\citep{1982ApJ...262..478G}.

@@

As the binary evolves, its increasing orbital period and speed, and
the apparant superluminal motion effect, together result in a jet
morphology with an increasing opening angle and increasingly rapid
periodic variation, together with a twisted shape.
Figure~\ref{fig:jet_mdot1.0} shows these three features for an
equal-mass binary with total mass $M=10^{10}$~M$_\odot$.  For
concreteness, we have assumed $i=40^\circ$ and $\beta=0.9$.  The
angular diameter distance to the binary is assumed to be $d =
100$~Mpc, which corresponds to redshift $z\sim 0.3$.  The forward jet
streched on the sky while the backward jet is compressed due to the
superluminal motion.  The right panel of Figure~\ref{fig:jet_mdot1.0}
zooms into the inner region of the forward jet and shows that the
binary creates wiggles in the jet morphology of the order of a
milli-arcsecond.  The duration of the binary evolution captured by
Figure~\ref{fig:jet_mdot1.0} is 100 yr.  Although the system can be
rescaled to different masses, times, and separation, the initial
separation in Figure~\ref{fig:jet_mdot1.0} is $\sim 10^{-3}$~pc, when
the binary enters it GW-dominated phase.

The increasing orbital speed of the binary leads to an increase in the
jet's precession angular velocity.  This increases the helicity of the
jet such that the jet is wound up increasingly tightly closer to the
source.  This gives rise to a ``chirping'' morphology as seen in
Figure~\ref{fig:jet_mdot1.0}.  The chirp is easier to resolve in jets
that are closer to the line of sight, where superluminal motion can
magnify the forward jet.  A chirping black hole is an unambiguous
signature of a rapidly evolving binary BH; the chirp is direct
reflection of the orbital velocity evolution in
Figure~\ref{fig:evolve}.  Together, the three features of the jet
morphology in Figure~\ref{fig:jet_mdot1.0}, viz., conical jet geometry
with an increasing opening angle, a radio chirp geometry, and an
asymmetrical shape, directly probe the dynamics of the binary in its
GW phase. Observations of this milliarcsecond geometry can thus
constrain GR in the strong field limit.

@@

These milli-arcsecond features can be resolved using long baseline
radio interferometry \citep{2001ChJAA...1..236Q}.  At higher
redshifts, jets are fainter and the wiggles become more difficult to
resolve.  However, it might be possible to resolve binaries with very
unequal mass ratios, eccentric orbits, or suitable inclinations.  The
future Square Kilometer Array (SKA) should make microarcsecond
resolution possible, which could increase the detection rate of these
objects \citep{2014arXiv1412.5971P}.

The jet morphology acts as an electromagnetic counterpart for chirping
gravitational wave sources.  BH binaries are expected to be the
brightest GW sources in the universe.  Current ground based GW
detectors such as LIGO are sensitive to low-mass BH binaries at high
redshift, which would be difficult to detect with this method.
However, future detectors such as eLISA are sensitive to GW emission
from $10^7$--$10^9$ M$_\odot$ BHs \citep{2013CQGra..30x4009S}.  Pulsar
Timing Arrays are sensitive to even higher binary masses.  Monitoring
of jet morphologies with features outlined in this paper will put a
lower bound on the expected abundance of bright GW sources for this
experiments. Moreover, combined GW and electromagnetic observations of
compact binary mergers should enable detailed studies of astrophysical
processes in the strong-field gravity regime
\citep{2005ApJ...629...15H}.

We make the code implementing the model presented in this Letter
available at \url{https://github.com/gkulkarni/JetMorphology} under
the MIT open source software license.  This code can be used to
explore the parameter space of the model.

@@

\bibliographystyle{mnras}
\bibliography{refs}

\bsp
\label{lastpage}
\end{document}
